---
title: "QTL Mapping with new dose values"
author: "Daniel Gatti"
date: "11/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(broom)
library(qtl2)

base_dir = '/media/dmgatti/hdb/projects/TB'
```

Gillian sent me the actual CFU values for each mouse in each batch. I had been using the previously reported values, but would like to try using the median dose.

I was surprised when mapping that using dose as a numeric covariate performed worse (gave lower LOD scores for know positives like s100a8 on chr 3) than using dose (or generation) as a factor. If the calculated doses are off, then maybe this explains that behavior.

Read in the phenotypes and new doses.

```{r read_pheno}
pheno = readr::read_csv(file.path(base_dir, 'data', 'phenotypes', 'tufts_do_tb_pheno_cleaned.csv'))
doses = readr::read_csv(file.path(base_dir, 'data', 'phenotypes', 'tufts_tb_batch_doses.csv'))
```

Look at the distribution of doses.

```{r dist_dose}
doses = doses %>%
          pivot_longer(cols = everything(), names_to = 'batch', values_to = 'cfu')

dose_mean = doses %>% 
              group_by(batch) %>% 
              summarize(mean_cfu   = mean(cfu, na.rm = TRUE),
                        median_cfu = median(cfu, na.rm = TRUE))

doses %>%
  ggplot() +
    geom_boxplot(aes(batch, cfu)) +
    geom_point(mapping = aes(batch, mean_cfu), data = dose_mean, color = 'red', size = 2)
```
The means and medians are a bit off for the higher groups. I want to see if they are log-normally distributed (by eye).

```{r dist_dose1,fig.height=6}
doses %>% 
  ggplot() +
    geom_histogram(aes(cfu)) +
    facet_wrap(~batch, ncol = 1)
```

Maybe we should take the geometric mean on these.

```{r dose_geom_mean}
doses = doses %>% 
          mutate(log_cfu = log(cfu))
dose_mean = dose_mean %>% 
              full_join(doses %>% 
                          group_by(batch) %>% 
                          summarize(geom_cfu = exp(mean(log_cfu, na.rm = TRUE))))
```
Plot the old means versus the geometric means.

```{r plot_mean_geom}
dose_mean %>% 
  ggplot(aes(mean_cfu, geom_cfu)) +
    geom_point() +
    geom_smooth(method = 'lm')+
    geom_abline(aes(slope = 1, intercept = 0))
```

Unsurprisingly, the arithmetic means overestimate the CFU values a bit. I'm not sure if this is enough to affect the QTL mapping. 

Get the batch values from the phenotype file. At this point, I don't have values for GB07.7 & 8.

```{r get pheno_dose}
pheno_cfu = pheno %>% 
              select(batch = expr, pheno_cfu = mtb_dose) %>% 
              distinct() %>% 
              filter(pheno_cfu > 0)
dose_mean = full_join(dose_mean, pheno_cfu)
```

Plot the Mtb CFU values from the phenotype file versus the geometric mean.

```{r plot_pheno_mean_geom}
dose_mean %>% 
  ggplot(aes(pheno_cfu, geom_cfu)) +
    geom_point() +
    geom_smooth(method = 'lm')+
    geom_abline(aes(slope = 1, intercept = 0))
```

Plot phenotypes using the new Mtb CFU doses.

```{r pheno_plot}
# This is the slow way. I couldn't find a stringr equivalent of match().
pheno %>% 
  mutate(new_dose = 0,
         new_dose = if_else(str_detect(expr, dose_mean$batch[1]), dose_mean$geom_cfu[1], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[2]), dose_mean$geom_cfu[2], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[3]), dose_mean$geom_cfu[3], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[4]), dose_mean$geom_cfu[4], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[5]), dose_mean$geom_cfu[5], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[6]), dose_mean$geom_cfu[6], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[7]), dose_mean$geom_cfu[7], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[8]), dose_mean$geom_cfu[8], new_dose),
         new_dose = if_else(str_detect(expr, dose_mean$batch[9]), dose_mean$geom_cfu[9], new_dose), 
         mouse = as.character(mouse)) %>%  
  select(mouse, gen, expr, new_dose, pct_wt_loss, cxcl5:s100a8) %>%
  pivot_longer(cols = pct_wt_loss:s100a8, names_to = 'phenotype', values_to = 'value') %>% 
  mutate(value = value + 1) %>%
  ggplot(aes(new_dose, value, group = new_dose)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_wrap(~phenotype)
```

See if linear models fit better using the original doses or the new doses.

```{r lm_fit_doses}
lm_orig = pheno %>% 
            select(mouse, mtb_dose, euth_day, pct_wt_loss:s100a8) %>% 
            pivot_longer(cols = pct_wt_loss:s100a8, names_to = 'phenotype', values_to = 'value') %>% 
            group_by(phenotype) %>% 
            nest() %>% 
            mutate(mod = map(data, ~ lm(value ~ mtb_dose + euth_day, data = .x)),
                   summ = map(mod, glance)) %>% 
            select(phenotype, summ) %>% 
            unnest(cols = c(summ))
```

```{r lm_fit_doses2}
lm_new = pheno %>% 
           mutate(new_dose = 0,
                  new_dose = if_else(str_detect(expr, dose_mean$batch[1]), dose_mean$geom_cfu[1], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[2]), dose_mean$geom_cfu[2], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[3]), dose_mean$geom_cfu[3], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[4]), dose_mean$geom_cfu[4], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[5]), dose_mean$geom_cfu[5], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[6]), dose_mean$geom_cfu[6], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[7]), dose_mean$geom_cfu[7], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[8]), dose_mean$geom_cfu[8], new_dose),
                  new_dose = if_else(str_detect(expr, dose_mean$batch[9]), dose_mean$geom_cfu[9], new_dose), 
                  mouse = as.character(mouse)) %>% 
            select(mouse, new_dose, euth_day, pct_wt_loss:s100a8) %>% 
            pivot_longer(cols = pct_wt_loss:s100a8, names_to = 'phenotype', values_to = 'value') %>% 
            group_by(phenotype) %>% 
            nest() %>% 
            mutate(mod = map(data, ~ lm(value ~ new_dose + euth_day, data = .x)),
                   summ = map(mod, glance)) %>% 
            select(phenotype, summ) %>% 
            unnest(cols = c(summ))
```

```{r}
plot(lm_orig$adj.r.squared, lm_new$adj.r.squared)
abline(0, 1)
```

The adjusted R-squared is lower when using the new Mtb doses (!!). 




