---
title: "Tufts Affy Array Data"
author: "Daniel Gatti"
date: "July 7, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)

#General Bioconductor packages
    library(Biobase)
    library(oligoClasses)
     
#Annotation and data import packages
    library(ArrayExpress)
    library(pd.mogene.2.0.st)
    library(mogene20sttranscriptcluster.db)
     
#Quality control and pre-processing packages
    library(oligo)
    library(arrayQualityMetrics)
     
#Analysis and statistics packages
    library(limma)
    library(topGO)
    library(ReactomePA)
    library(clusterProfiler)
    library(pcaMethods)
     
#Plotting and color options packages
    library(gplots)
    library(ggplot2)
    library(geneplotter)
    library(RColorBrewer)
    library(pheatmap)
     
#Formatting/documentation packages
    library(rmarkdown)
    library(dplyr)
    library(tidyr)

#Helpers:
    library(stringr)
    library(matrixStats)
    library(genefilter)
    library(openxlsx)
    library(devtools)

base_dir   = '/media/dmgatti/hdb/projects/TB'
cel_dir    = file.path(base_dir, 'data', 'expression', 'cel_files', 'analysis', 'raw')
report_dir = file.path(base_dir, 'data', 'expression', 'cel_files', 'analysis', 'report')
sdrf_file  = file.path(cel_dir, 'tufts_do_tb_sdrf.tsv')
```

```{r read_cel_files}
SDRF = read.delim(sdrf_file)
rownames(SDRF) = SDRF$Array.Data.File
SDRF = AnnotatedDataFrame(SDRF)

raw_data = oligo::read.celfiles(filenames = file.path(cel_dir, SDRF$Array.Data.File),
                                verbose = FALSE, phenoData = SDRF)
stopifnot(validObject(raw_data))
```


```{r pca_raw}
exp_raw = log2(Biobase::exprs(raw_data))
PCA_raw = prcomp(t(exp_raw), scale. = TRUE)

tmp = data.frame(PC1 = PCA_raw$x[,1], PC2 = PCA_raw$x[,2], PC3 = PCA_raw$x[,3], 
                 batch = as.character(pData(raw_data)$Factor.Value.batch),
                 expt = pData(raw_data)$Factor.Value.expt,
                 dose = as.factor(pData(raw_data)$Factor.Value.dose))

ggplot(tmp, aes(PC1, PC2, color = batch)) +
  geom_point(size = 2)

ggplot(tmp, aes(PC2, PC3, color = batch)) +
  geom_point(size = 2)
# Array batch shows up in PC3.
```

```{r pca_raw_dose}
ggplot(tmp, aes(PC2, PC3, color = dose)) +
  geom_point(size = 2)
```


```{r qaqc}
oligo::boxplot(raw_data, target = "core", 
               main = "Boxplot of log2-intensitites for the raw data")

arrayQualityMetrics(expressionset = raw_data,
                    outdir = report_dir,
                    force = TRUE, do.logtransform = TRUE,
                    intgroup = c("Factor.Value.batch", "Factor.Value.dose"))
```


```{r rma}
# RMA
rma_eset = oligo::rma(raw_data, target = "core", normalize = FALSE)
```

```{r rle}
# RLE
row_medians_assayData = Biobase::rowMedians(as.matrix(Biobase::exprs(rma_eset)))

RLE_data = sweep(Biobase::exprs(rma_eset), 1, row_medians_assayData)

RLE_data = as.data.frame(RLE_data)
RLE_data_gathered = tidyr::gather(RLE_data, array, log2_expression_deviation)

ggplot(RLE_data_gathered, aes(array, log2_expression_deviation)) + 
  geom_boxplot(outlier.shape = NA) + 
  ylim(c(-3, 3)) + 
  theme(axis.text.x = element_text(colour = "aquamarine4", 
                                   angle = 60, size = 6.5, hjust = 1 ,
                                   face = "bold"))
```

I'm not removing any arrays for now. The low expression ones could be the tail of the distribution.

```{r rma_core}
# RMA
rma_eset = oligo::rma(raw_data, target = "core")
```

```{r pca_rma_core}
expr = Biobase::exprs(rma_eset)
PCA  = prcomp(t(expr), scale = TRUE)

percentVar = round(100*PCA$sdev^2/sum(PCA$sdev^2),1)
sd_ratio   = sqrt(percentVar[2] / percentVar[1])

dataGG = data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2], PC3 = PCA$x[,3],
                    batch = factor(Biobase::pData(rma_eset)$Factor.Value.batch),
                    dose  = factor(Biobase::pData(rma_eset)$Factor.Value.dose))

ggplot(dataGG, aes(PC1, PC2)) +
      geom_point(aes(shape = batch, colour = dose)) +
  ggtitle("PCA plot of the calibrated, summarized data") +
  xlab(paste0("PC1, VarExp: ", percentVar[1], "%")) +
  ylab(paste0("PC2, VarExp: ", percentVar[2], "%")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_shape_manual(values = c(4,15)) + 
  scale_color_manual(values = c("darkorange2", "dodgerblue4", 'plum4'))
```

```{r}
ggplot(dataGG, aes(PC2, PC3)) +
      geom_point(aes(shape = batch, colour = dose)) +
  ggtitle("PCA plot of the calibrated, summarized data") +
  xlab(paste0("PC2, VarExp: ", percentVar[2], "%")) +
  ylab(paste0("PC3, VarExp: ", percentVar[3], "%")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_shape_manual(values = c(4,15)) + 
  scale_color_manual(values = c("darkorange2", "dodgerblue4", 'plum4'))
```


```{r rma_heatmap,fig.width=12,fig.height=12}
# Heatmap
expr = exprs(rma_eset)

dists = as.matrix(dist(t(expr), method = "manhattan"))

rownames(dists) <- row.names(pData(rma_eset))
hmcol <- rev(colorRampPalette(RColorBrewer::brewer.pal(9, "YlOrRd"))(255))
colnames(dists) <- NULL
diag(dists) <- NA

annotation_for_heatmap = data.frame(batch = pData(rma_eset)$Factor.Value.batch,  
                                    dose  = pData(rma_eset)$Factor.Value.dose)

row.names(annotation_for_heatmap) <- row.names(pData(rma_eset))

ann_colors <- list(
  batch = c('2017' = "chartreuse4", '2019' = "burlywood3"),
  dose  = c('0' = "blue4", '16' = "cadetblue2", '28' = 'royalblue')
)

pheatmap(dists, col = (hmcol), 
         annotation_row = annotation_for_heatmap,
         annotation_colors = ann_colors,
         legend = TRUE, 
         treeheight_row = 0,
         legend_breaks = c(min(dists, na.rm = TRUE), 
                           max(dists, na.rm = TRUE)), 
         legend_labels = (c("small distance", "large distance")),
         main = "Clustering heatmap for the calibrated samples")
```

Samples 307 & 313 are further from the other samples. Still leaving them in for now.

```{r filtering}
expr_medians = rowMedians(Biobase::exprs(rma_eset))

hist_res = hist(expr_medians, 100, col = "cornsilk1", freq = FALSE, 
            main = "Histogram of the median intensities", 
            border = "antiquewhite4",
            xlab = "Median intensities")
```

I'm keeping all genes becuase I will perform ComBat batch normalization next.

```{r write_expr}
write.csv(Biobase::exprs(rma_eset), file = file.path(base_dir, 'data', 'expression', 'do_tb_expr_rma.csv'),
          quote = F)
```

Combat batch normalization.

```{r combat}
mod   = model.matrix(~Factor.Value.dose, data = pData(rma_eset))
batch = as.character(pData(rma_eset)$Factor.Value.batch)
expr_cb = sva::ComBat(dat = exprs(rma_eset), batch = batch, mod = mod)
```

Check PCA after ComBat.

```{r pca_post_combat}
PCA  = prcomp(t(expr_cb), scale = TRUE)

percentVar = round(100*PCA$sdev^2 / sum(PCA$sdev^2),1)
sd_ratio   = sqrt(percentVar[2] / percentVar[1])

dataGG = data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2], PC3 = PCA$x[,3],
                    batch = factor(Biobase::pData(rma_eset)$Factor.Value.batch),
                    dose  = factor(Biobase::pData(rma_eset)$Factor.Value.dose))

ggplot(dataGG, aes(PC1, PC2)) +
      geom_point(aes(shape = batch, colour = dose)) +
  ggtitle("PCA plot of the calibrated, summarized data") +
  xlab(paste0("PC1, VarExp: ", percentVar[1], "%")) +
  ylab(paste0("PC2, VarExp: ", percentVar[2], "%")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_shape_manual(values = c(4,15)) + 
  scale_color_manual(values = c("darkorange2", "dodgerblue4", 'plum4'))
```

```{r pca_post_combat_pc23}
ggplot(dataGG, aes(PC2, PC3)) +
      geom_point(aes(shape = batch, colour = dose)) +
  ggtitle("PCA plot of the calibrated, summarized data") +
  xlab(paste0("PC2, VarExp: ", percentVar[2], "%")) +
  ylab(paste0("PC3, VarExp: ", percentVar[3], "%")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_shape_manual(values = c(4,15)) + 
  scale_color_manual(values = c("darkorange2", "dodgerblue4", 'plum4'))
```

Add combat expression back to eset.

```{r write_expr_cb}
write.csv(expr_cb, file = file.path(base_dir, 'data', 'expression', 'do_tb_expr_rma_combat.csv'),
          quote = F)
exprs(rma_eset) = expr_cb
```

Add transcript cluster mappings.

```{r add_transcr_cl}
annot = AnnotationDbi::select(mogene20sttranscriptcluster.db,
                              keys = featureNames(rma_eset),
                              columns = c('ENSEMBL', 'ENTREZID', 'SYMBOL', 'GENENAME'),
                              keytype = 'PROBEID')
# Remove transcripts with no Ensembl mapping (??)
annot = subset(annot, !is.na(ENSEMBL))
```

Keep probesets that only map to one gene (for now).

```{r}
annot_keep = dplyr::count(annot, PROBEID) %>%
              filter(n == 1)
annot = subset(annot, PROBEID %in% pull(annot_dupl, PROBEID))

rma_eset = subset(rma_eset, featureNames(rma_eset) %in% pull(annot, PROBEID))

fData(rma_eset)$PROBEID = rownames(fData(rma_eset))
fData(rma_eset) = left_join(fData(rma_eset), annot)
```

```{r}
saveRDS(rma_eset, file = file.path(base_dir, 'data', 'expression', 'do_tb_expr_rma_combat_eset.rds'))
```



