---
title: 'Tufts DO TB: Gather qtl2 data'
author: "Daniel Gatti"
date: "February 22, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(readxl)
library(tidyverse)
library(qtl2)
library(qtl2convert)

base_dir = '/media/dmgatti/hdb/projects/TB'
pheno_file  = file.path(base_dir, 'data', 'phenotypes', 'tufts_do_tb_pheno_cleaned.csv')
fda_pheno_file = '/media/dmgatti/hdb/projects/USFDA/data/phenotypes/fda_pheno_cleaned.csv'
image_file  = file.path(base_dir, 'data', 'image_measurements.xlsx')
probs_file  = file.path(base_dir, 'haplo_reconstr', 'tufts_do_alleleprobs_cleaned.rds')
fda_probs_file = '/media/dmgatti/hdb/projects/USFDA/haplo_reconstr/haps/fda_do_alleleprobs_qc_filtered.rds'
marker_file = file.path(base_dir, 'haplo_reconstr', 'qtl2', 'pmap.csv')
output_file = file.path(base_dir, 'data', 'tufts_do_tb_qtl2_input.Rdata')
```

Load in phenotypes, genoprobs & markers. Markers are from U. Wisc. coordinates.

```{r read_data}
pheno     = readr::read_csv(pheno_file)
fda_pheno = readr::read_csv(fda_pheno_file)
img_data  = read_xlsx(path = image_file)
colnames(img_data) = c('mouse', 'prSS', 'necr')
probs     = readRDS(probs_file)
fda_probs = readRDS(fda_probs_file)
markers   = readr::read_csv(marker_file, col_types = 'ccn')
```

Align samples between pheno and probs.

```{r align_samples}
# Keep only DO mice dosed with TB.
pheno = filter(pheno, strain == 'J:DO' & mtb_dose > 0)

pheno = full_join(pheno, img_data)

# Align samples between phenotypes and probs.
samples = intersect(pheno$mouse, rownames(probs[[1]]))
pheno = as.data.frame(pheno)
pheno$mouse = as.character(pheno$mouse)
rownames(pheno) = pheno$mouse
pheno = pheno[samples,]
probs = probs[samples,]

stopifnot(all(nrow(pheno) == nrow(probs[[1]])))
stopifnot(all(rownames(pheno) == rownames(probs[[1]])))
```

There are `r nrow(pheno)` samples in common between the phenotypes and genoprobs.

Combine the FDA mouse number, dose and survival data for non-vaccinted mice with the Tufts data.

```{r merge_fda_tufts}
fda_pheno = fda_pheno %>%
              filter(vacc == 'N') %>%
              select(mouse = sample, gen, sex, vacc, dose, survival, cause, lung_mtb)
fda_pheno = as.data.frame(fda_pheno)
rownames(fda_pheno) = fda_pheno$mouse
common = intersect(fda_pheno$mouse, rownames(fda_probs[[1]]))

fda_pheno = fda_pheno[common,]
fda_probs = fda_probs[common,]

pheno = full_join(pheno, fda_pheno)
names(fda_probs) = sub('^chr', '', fda_probs)
probs = rbind(probs, fda_probs)
```


Separate phenotypes from covariates.

```{r make_pheno_covar}
covar = select(pheno, mouse:euth)
rownames(covar) = covar$mouse
pheno = select(pheno, mouse, pct_wt_loss:resistin, necr)
```

RankZ transform the phenotypes.

```{r log_rankz_pheno}
rankZ = function(x) {
  qnorm((rank(x, na.last = 'keep') - 0.5 ) / sum(!is.na(x)))
} # rankZ()

rownames(pheno) = pheno$mouse
pheno = as.matrix(pheno[,-1])
pheno = apply(pheno, 2, rankZ)
```

Turn covariates into factors. I'm not sure about Mtb_dose yet. Last time, adding it as a factor was better than adding it as a numeric covariate.

```{r covar_factor}
covar$gen  = factor(covar$gen)
covar$expr = factor(covar$expr)
covar$cage = factor(covar$cage)
covar$aero_run = factor(covar$aero_run)
covar$euth = factor(covar$euth)
```

Align markers between markers and probs.

```{r align_markers}
markers = as.data.frame(markers)
rownames(markers) = markers$marker
map = qtl2convert::map_df_to_list(markers, pos_column = 'pos')

unique_chr = names(probs)

for(chr in unique_chr) {
  
  mkr = intersect(names(map[[chr]]), dimnames(probs[[chr]])[[3]])

  map[[chr]]  = map[[chr]][mkr]
  probs[[chr]] = probs[[chr]][,,mkr]
  stopifnot(all(names(map[[chr]]) == dimnames(probs[[chr]])[[3]]))

} # for(chr)
```

Create kinship matrices.

```{r make_kinship}
K = calc_kinship(probs, type = 'loco', cores = 4)
```

Save large Rdata file.

```{r save_data}
save(pheno, covar, map, probs, K, file = output_file)
```













